---
title: "Washington PM2.5 and Heat Episides, 2010-2018"
output: github_document
---

```{r setup, include=F}
knitr::opts_chunk$set(echo = TRUE, fig.width=12, fig.height=8, warning=FALSE, message=FALSE)

```

```{r libraries, include=F}

library(pacman)


```

```{r load.data.themes, include=F}

source("Data and Figure Format.R")

```

```{r heat-index-function, include=F}

source("Heat Index Function.R")

```


## METHODS

### Heat

- Requested AWN data (with data QA date pre-2010) from locations nearest EPA PM sites for 2010-2018
- Computed heat index using Rothfusz approach

### PM2.5

- Downloaded [EPA PM data for 2010-2018](#https://aqs.epa.gov/aqsweb/airdata/download_files.html)
- Merged with heat data
- Computed nearest Haversine distance between AWN and EPA sites

### Plots

- Histograms for temp, relative humidity, heat index, and PM2.5 (PM2.5 FRM/FEM Mass (88101))
- Density plots
- Time series
- Scatter plot of heat index vs. PM, 2010-2018 by location (n=24)

### Maps

- Ag worker population by county
- PM * HI by county

## RESULTS

```{r summary.tables}

n.am.asd.gm.gsd(epa.awn.final.wide$Sample.Measurement)
n.am.asd.gm.gsd(epa.awn.final.wide$hi)

```

```{r histograms.temp.rh.hi}

ggplot(epa.awn.final.wide, aes(x=AIR_TEMP_F))            + 
  geom_histogram() + 
  theme_DEOHS(10) + 
  xlab("Air Temperature (F)") + 
  ylab("Count")

ggplot(epa.awn.final.wide, aes(x=`RELATIVE_HUMIDITY_%`)) + 
  geom_histogram() + 
  theme_DEOHS(10) + 
  xlab("Relative Humidity (%)") + 
ggplot(epa.awn.final.wide, aes(x=hi)) + 
  geom_histogram() + 
  theme_DEOHS(10) + 
  xlab("Heat Index") + 
  ylab("Count")

asinh_trans <- function(){
  trans_new(name = 'asinh', transform = function(x) asinh(x), 
            inverse = function(x) sinh(x))
}

custom_log_x_trans <- function(){
  trans_new("custom_log_x",
            transform = function (x) ( sign(x)*log(abs(x)+1) ),
            inverse = function (y) ( sign(y)*( exp(abs(y))-1) ),
            domain = c(-Inf,Inf))
}

ggplot(epa.awn.final.wide, aes(x=Sample.Measurement)) + 
  geom_histogram() + 
  coord_trans(x = "custom_log_x") + 
  theme_DEOHS(10) + 
  xlab(expression(paste("P", M[2.5], " (", mu,"g/", m^3,")"))) +
  ylab("Count")

ggplot(epa.awn.final.wide, 
       aes(County.Name, Sample.Measurement))  + 
  geom_violin(fill = "grey") +
  coord_trans(y = "custom_log_x") + 
  geom_hline(yintercept = 35, color = "orange") +
  facet_wrap(~County.Name, scales = "free_x") +
  theme_DEOHS(10)+ 
  ylab(expression(paste("P", M[2.5], " (", mu,"g/", m^3,")"))) +
  xlab("County Name")


```

```{e PM25trend, include = F, cache = T}

ggplot(epa.all.1h.summary[, list(PM = mean(PM)), 
                          by = c("County","year")],
       aes(year, PM)) + 
  geom_point(aes(color = `County`)) +
  theme_DEOHS(10) + 
  geom_smooth(span = 1, se = T) +
  ylab(expression(paste("P", M[2.5], " (", mu,"g/", m^3,")"))) +
  xlab("Year")


```

```{r density-plots}

ggplot(epa.awn.final.wide[!is.na(hi)], aes( `hi`, y=..density..)) + 
  geom_density() + 
  geom_vline(xintercept = 80, col = "red") +
  facet_wrap(~County.Name, scales = "free") +
  geom_histogram(aes(y=..density..), alpha=0.5, position="identity") +
  xlim(0, 100) + 
  theme_DEOHS(12) + 
  xlab("Heat Index")

ggplot(epa.awn.final.wide[!is.na(`Sample.Measurement`)], 
       aes( `Sample.Measurement`, y=..density..)) + 
  geom_density() + 
  coord_trans(x = "sqrt") + 
  geom_vline(xintercept = 35, col = "red") +
  facet_wrap(~County.Name, scales = "free") +
  geom_histogram(aes(y=..density..), alpha=0.5, position="identity")+
  xlim(0, 100) + 
  theme_DEOHS(12) + 
  xlab(expression(paste("P", M[2.5], " (", mu,"g/", m^3,")"))) 

```

```{r time.series}

ggplot(epa.awn.final.wide, 
       aes(x=datetime_local, 
           y=Sample.Measurement, 
           color="PM2.5")) + 
  geom_line() + 
  geom_line(data = epa.awn.final.wide,
            aes(x=datetime_local, 
           y=hi, 
           color="Heat Index") ) +
  facet_wrap(~County.Name) + 
  theme_DEOHS(12) + 
  xlab("Year") + 
  ylab("")

#remove Garfield, Klickitat and Pend Oreille

epa.awn.final.wide = epa.awn.final.wide[!(County.Name %in% c("Garfield","Klickitat","Pend Oreille")),]

```

```{r scatter.plots}
#All data

ggplot(epa.awn.final.wide, 
       aes(x=hi, y=Sample.Measurement)) + 
  geom_point(size=1) + 
  facet_wrap(facets = vars(County.Name)) + 
  theme_DEOHS(12) + 
  xlab("Heat Index") + 
  ylab("PM2.5 µg/m3")

#PM data above 35 ug/m3

ggplot(epa.awn.final.wide[epa.awn.final.wide$Sample.Measurement>35], 
       aes(x=hi, y=Sample.Measurement)) + 
  geom_point(size=1) + 
  facet_wrap(~County.Name) + 
  theme_DEOHS(12) + 
  xlab("Heat Index") + 
  ylab("PM2.5 µg/m3")

#Heat index data above 80
#

#consider 85 threshold
ggplot(epa.awn.final.wide[epa.awn.final.wide$hi>85], 
       aes(x=hi, y=Sample.Measurement)) + 
  geom_point(size=1) + 
  facet_wrap(facets = vars(Site)) + 
  theme_DEOHS(12) + 
  xlab("Heat Index") + 
  ylab("PM2.5 µg/m3")

#density plot
p_load(viridis)
df <- fortify(epa.awn.final.wide[`Sample.Measurement`>=35 |
                          `hi`>=85])

#table: burden of workers impacted by combined heat and airborne particulate matter exposures
p_load(lubridate)
df[, quarter_val := quarter(hour)]
epa.awn.final.wide[, quarter_val := quarter(hour)]

above_threshold_county = epa.awn.final.wide[, list("Hours HI" = sum(hi>=85, na.rm=T)/length(hi)*100,
          "Hours PM2.5" = sum(Sample.Measurement >= 35, na.rm=T)/length(hi)*100 ,
          "Hours Combined" = sum(Sample.Measurement >= 35 & hi>=85, na.rm=T)/length(hi)*100),
   by = c("County.Name", "quarter_val")]

write.csv(dcast(above_threshold_county,   as.character(County.Name) ~ quarter_val, 
      value.var = c("Hours HI","Hours PM2.5", "Hours Combined" )), "Data/high_hours.csv")

#add naics descriptions
naics = fread("Data/2-6 digit_2017_Codes.csv")
naics = naics[substr(`2017 NAICS US   Code`, 1, 2) == "11"]
naics = naics[nchar(`2017 NAICS US   Code`)==4,1:3]
setkey(naics, `2017 NAICS US   Code`)
setkey(qwi_ag, industry)
qwi_ag = naics[qwi_ag]

#Add county names
p_load(dint)
p_load(tidycensus)
p_load(tidyverse)
p_load(magrittr)


fips = copy(fips_codes)
fips = setDT(fips)

fips = fips[state_code %in% "53",]
fips[, county := sub(" County", "", county)]
setkey(fips, county_code)
setkey(qwi_ag, county)
setkey(qwi_ag_all, county)
qwi_ag = qwi_ag[fips]
qwi_ag_all = qwi_ag_all[fips]
qwi_ag[, yrq := date_yq(year, quarter)]
qwi_ag_all[, yrq := date_yq(year, quarter)]
setnames(qwi_ag, "i.county", "County.Name")
setnames(qwi_ag_all, "i.county", "County.Name")

qwi_ag_all_summary = 
  qwi_ag_all[,list("Ag Workers" = sum(Emp)/length(unique(year))), 
             by = c("County.Name", "quarter")  ]

dcast(qwi_ag_all_summary,   County.Name ~ quarter, 
      value.var = c("Ag Workers"))


ggplot(epa.awn.final.wide[`Sample.Measurement`>=35 &
                          `hi`>=85]) +
  geom_hex(bins = 10, aes(x    = `Sample.Measurement`,
               y    = hi,
               fill = ..count..),
           color = "black") +
  scale_fill_viridis_c(option = "inferno") +
  labs(fill = 'Count') +
  facet_wrap(~County.Name) +
  xlab(expression(paste("P", M[2.5], " (", mu,"g/", m^3,")"))) +
  theme_DEOHS(10) +
  xlim(0,200) +
  geom_hline(yintercept = 85, col = "red")+
  geom_vline(xintercept = 35, col = "red") +
  theme(legend.key.size = unit(3,"line"))

epa.awn.final.wide[, yrq := date_yq(year.x, quarter_val)]

above_threshold_county_hours = 
  epa.awn.final.wide[, list("Hours HI" = sum(hi>=85, na.rm=T)/length(hi)*2190,
          "Hours PM2.5" = sum(Sample.Measurement >= 35, na.rm=T)/length(hi)*2190 ,
          "Hours Combined" = sum(Sample.Measurement >= 35 & hi>=85, na.rm=T)/length(hi)*2190),
   by = c("County.Name", "quarter_val")]


qwi_ag_sum = qwi_ag_all[, list(EmpAll = mean(Emp)), by = c("County.Name","quarter")]
#hours above threshold 
above_threshold_county_hours[, County.Name := as.factor(County.Name)]
qwi_ag_sum[, County.Name := as.factor(County.Name)]

qwi_ag_sum$County.Name <- reorder(qwi_ag_sum$County.Name, qwi_ag_sum$EmpAll)
above_threshold_county_hours$County.Name <- 
  factor(above_threshold_county_hours$County.Name, 
          levels = levels(qwi_ag_sum$County.Name))

setkey(above_threshold_county_hours, County.Name, quarter_val)
setkey(qwi_ag_sum, EmpAll)

#Figure 1 Employment vs heat/PM
ggplot(above_threshold_county_hours) +
  geom_line(aes(quarter_val, `Hours HI`, color = "Hours of Heat exposure"), size = 2) +
  geom_line(aes(quarter_val, `Hours PM2.5`, color = "Hours of PM exposure"), size = 2) +
  geom_line(aes(quarter_val, `Hours Combined`, color = "Hours of Combined exposure"), size = 2) +
  scale_colour_Publication(name = "")+ 
  geom_bar(data = qwi_ag_sum, 
         aes(quarter, EmpAll*350/35000, fill = "Agricultural Employment"), 
         stat="identity", alpha = 0.2) +
  facet_wrap(~County.Name) +
  xlab("Quarter of the Year") +
  scale_x_continuous(labels = function(b) { paste0("Q", b)}) +
  theme_DEOHS(12) + 
  scale_fill_manual(name = "", values = "darkgrey") +
  scale_y_continuous(
    trans = "sqrt",
    name = "Number of Hours (h)", 
    sec.axis = sec_axis(~ . * 35000 / 350, name = "Number of Agricultural Workers",
                     labels = function(b) { paste0(round(b/1000, 0), "K")})) +
  theme(legend.key.size = unit(3,"line")) +
  facet_wrap(~County.Name) 

#Time Series for Yakima
epa.awn.yakima = epa.awn.final.wide[County.Name == "Yakima",]

#Figure 2, Time Series Yakima
ggplot(epa.awn.yakima[`yrq` %in% date_yq(c(2015:2018), 3),]) +
  geom_point(aes(x = datetime_local, y = hi, color = "Heat Index"), alpha = 0.2) + 
  geom_point(aes(x = datetime_local, 
                 y = Sample.Measurement, color = "PM Concentration"), alpha = 0.2) +
  ggtitle("Yakima County, 3rd Quarter") + 
  theme_DEOHS(12) +
  scale_colour_Publication(name = "") + 
   scale_y_continuous(sec.axis = sec_axis(~ . , name = "Particulate Matter PM2.5")) +
  ylab("Heat Index (F)") + xlab("") +
  geom_hline(yintercept = 85 , color = "#386cb0", size = 1.5 ) +
  geom_hline(yintercept = 35 , color = "#fdb462", size = 1.5 ) +
  theme(axis.title.y = element_text(color="#386cb0"),
      axis.title.y.right = element_text(color= "#fdb462")) +
  guides(color = guide_legend(override.aes = list(size=5))) +
  facet_wrap(~year.x, scales = "free_x", nrow = 1)

epa.awn.okanogan = epa.awn.final.wide[County.Name == "Okanogan",]

#Figure 2, Time Series Okanogan
ggplot(epa.awn.okanogan[`yrq` %in% date_yq(c(2015:2018), 3),]) +
  geom_point(aes(x = datetime_local, y = hi, color = "Heat Index"), alpha = 0.2) + 
  geom_point(aes(x = datetime_local, 
                 y = Sample.Measurement, color = "PM Concentration"), alpha = 0.2) +
  theme_DEOHS(12) +
  scale_colour_Publication(name = "") + 
  ggtitle("Okanogan County, 3rd Quarter") + 
   scale_y_continuous(
     limits = c(0,350),
    sec.axis = sec_axis(~ . , name = "Particulate Matter PM2.5")) +
  ylab("Heat Index (F)") + xlab("") +
  geom_hline(yintercept = 85 , color = "#386cb0", size = 1.5 ) +
  geom_hline(yintercept = 35 , color = "#fdb462", size = 1.5 ) +
  theme(axis.title.y = element_text(color="#386cb0"),
      axis.title.y.right = element_text(color= "#fdb462")) +
  guides(color = guide_legend(override.aes = list(size=5))) +
  facet_wrap(~year.x, scales = "free_x", nrow = 1)

qwi_ag[, yrq := date_yq(year, quarter)]

#Figure Seasonal Employment Agriculture
qwi_ag$County.Name <- 
  factor(qwi_ag$County.Name, 
          levels = levels(qwi_ag_sum$County.Name))

ggplot(qwi_ag[County.Name %in% levels(qwi_ag$County.Name)[31:39]], 
       aes(yrq, Emp, color = `2017 NAICS US Title`)) + 
  geom_line(size = 1.4) +
  facet_wrap(~County.Name) +
  scale_colour_viridis_d(option = "C")+
  theme_DEOHS(12)

df[, yrq := date_yq(year.y, quarter_val)]

total_excess = df[, length(hour), by = c("yrq","County.Name")]

setkey(total_excess, yrq, County.Name)
setkey(qwi_ag_res, yrq, i.county )

qwi_ag_res[, Emp := sum(Emp, na.rm=T), by = c("yrq","i.county")]

combined = total_excess[qwi_ag_res]

combined = combined[!is.na(V1)]

summary_list = combined[, list(Exceedances = (V1),
                Employees = (Emp)), by = c("yrq","County.Name")]

setkey(summary_list, County.Name, yrq)

ggplot(summary_list, aes(yrq, Exceedances, color =  Employees)) + geom_point(size = 3) + 
  facet_wrap(~County.Name)  + 
  stat_smooth()+
    scale_color_viridis("test",trans = "log10") + theme_DEOHS(12) +
  ylab("Combined Heat and PM Hours") +
  theme(legend.position = "right", legend.direction="vertical") 



```

## REFERENCES
