---
title: "Washington PM2.5 and Heat Episides, 2010-2018"
output: github_document
---

```{r setup, include=F}
knitr::opts_chunk$set(echo = TRUE, fig.width=12, fig.height=8, warning=FALSE, message=FALSE)

```

```{r libraries, include=F}

library(pacman)
p_load(plyr)
p_load(readr)
p_load(data.table)
p_load(lubridate)
p_load(ggplot2)  
p_load(scales)   
p_load(gridExtra) 
p_load(grid) 
p_load(dplyr)
p_load(xts)
p_load(sqldf)
p_load(rgeos)
p_load(geosphere)
p_load(tidyr)
p_load(tidyqwi)
p_load(httr)
p_load(sp)

```

```{r load.theme, include=F}

#### Koundy's theme ####
# Source: https://github.com/koundy/ggplot_theme_Publication/blob/master/R/ggplot_theme_Publication.R
theme_HEI <- function(base_size=18) {
  library(grid)
  library(ggthemes)
  (theme_foundation(base_size=base_size)
  + theme(plot.title = element_text(face = "bold",
                                    size = rel(1.2), hjust = 0.5),
          text = element_text(),
          panel.background = element_rect(colour = NA),
          plot.background = element_rect(colour = NA),
          panel.border = element_rect(colour = NA),
          axis.title = element_text(face = "bold",size = rel(1)),
          axis.title.y = element_text(angle=90,vjust =2),
          axis.title.x = element_text(vjust = -0.2),
          axis.text = element_text(), 
          axis.line = element_line(colour="black"),
          axis.ticks = element_line(),
          panel.grid.major = element_line(colour="#f0f0f0"),
          panel.grid.minor = element_blank(),
          legend.key = element_rect(colour = NA),
          legend.position = "bottom",
          legend.direction = "horizontal",
          legend.key.size= unit(0.5, "cm"),
          legend.margin = unit(0, "cm"),
          legend.title = element_blank(),
          plot.margin=unit(c(4,4,4,4),"mm"),
          strip.background=element_rect(colour="#f0f0f0",fill="#f0f0f0"),
          strip.text = element_text(face="bold")
  ))
  
}

```

```{r heat-index-function, include=F}

#create heat index function
heat_index <- function(TempF, RH, units = "F") {
  #Based on the heat index equation: 
  #https://www.wpc.ncep.noaa.gov/html/heatindex_equation.shtml
  #Rothfusz approach
  
  datarh  = data.table(TempF, RH)
  
  if (units == "C"){
    datarh[, TempF := (TempF * 9/5) + 32]
  }
  
  datarh[, HI := 0.5 * (TempF + 61.0 + ((TempF-68.0)*1.2) + (RH*0.094))]
  
  datarh[, HIave := rowMeans(.SD), .SDcols = c("HI", "TempF")]
  
  datarh[HIave > 80, HI := -42.379 + 2.04901523*TempF + 10.14333127*RH - .22475541*TempF*RH - 
           .00683783*TempF*TempF - 
           .05481717*RH*RH + .00122874*TempF*TempF*RH + 
           .00085282*TempF*RH*RH - .00000199*TempF*TempF*RH*RH ]
  
  datarh$ADJUSTMENT = 0
  datarh[RH < 13 & (TempF>=80 & TempF <= 112), 
    ADJUSTMENT := -((13-RH)/4)*sqrt((17-abs(TempF-95.))/17)]
  datarh[RH > 85 & (TempF>=80 & TempF <= 87), 
    ADJUSTMENT := ((RH-85)/10) * ((87-TempF)/5)]
  
  datarh = datarh[, HI := HI + ADJUSTMENT]
  datarh$HI
    
}

```

```{r am-asd-gm-gsd-function, include=F}
# Write and use GM, GSD, AM, ASD functions

## No error handling

gm<-function(x){
    rslt<-exp(mean(log(x))) 
    return(rslt)
    }

gsd<-function(x){
    rslt<-exp(sqrt(var(log(x))))
    return(rslt)
    }

## Error handling

gm_eh<-function(x){
    if(sum(is.na(x))>0)
    {return(NA)}
    if(sum(!is.finite(log(x)))==0)
    {rslt<-exp(mean(log(x)))} # problem if x=0 -> log
    else{rslt<-Inf}
    return(rslt)
    }

gsd_eh<-function(x){
    if(sum(is.na(x))>0)
    {return(NA)}
    if(sum(!is.finite(log(x)))==0)
    {rslt<-exp(sqrt(var(log(x))))} # problem if x=0 -> log
    else{rslt<-Inf}
    return(rslt)
    }

n.am.asd.gm.gsd<-function(dta){
    x<-subset(dta, !is.na(dta))
    return(c( N=length(x),AM=mean(x),ASD=sd(x),GM=gm_eh(x),GSD=gsd_eh(x), Min=min(x), Max=max(x) ))
    }

```

```{r get-equation-rsquared-function, include=F}

lm_eqn = function(m) {

  l <- list(a = format(coef(m)[1], digits = 2),
      b = format(abs(coef(m)[2]), digits = 2),
      r2 = format(summary(m)$r.squared, digits = 3));

  if (coef(m)[2] >= 0)  {
    eq <- substitute(italic(y) == a + b %.% italic(x)*","~~italic(r)^2~"="~r2,l)
  } else {
    eq <- substitute(italic(y) == a - b %.% italic(x)*","~~italic(r)^2~"="~r2,l)    
  }

  as.character(as.expression(eq));                 
}

```

```{r get_qwi_data, include = F, cache = T}

#query qwi employment data
# apikey = as.character(fread("API key/QWI key", header = F))
# qwi = get_qwi(years = as.character(2010:2019), 
#               quarters = c(1, 2, 3, 4), 
#               industry_level = "4", 
#               states = "53",
#               variables = c("sEmp", "Emp"),
#               endpoint = "rh", 
#               geography = "county", 
#               seasonadj = "U", 
#               apikey = apikey,
#               processing = "sequential")
#saveRDS(qwi, "Data/qwi_2010_2019_RH.RDS")
#
# qwi = readRDS("Data/qwi_2010_2019_RH.RDS")
# qwi_ag = setDT(qwi)
# 
# qwi_ag = qwi_ag[, NAICS2 := substr(qwi$industry,1,2) ]
# qwi_ag  = qwi_ag[NAICS2 %in% "11"]
# qwi_ag = add_qwi_labels(qwi_ag)
# 
# save(qwi_ag, file = "Data/qwi_ag_2010_2019_RH.RData")

```


```{r upload-clean-awn-data, include=F, cache=T}

# awn <- list.files("Data/awn", pattern=".csv", full.names = TRUE) %>%  lapply(read_csv) %>% bind_rows 
# 
# awn.temp.rh.na  <- subset(awn, select = c(TSTAMP_PST, UNIT_ID, STATION_NAME, LATITUDE, LONGITUDE, AIR_TEMP_F, `RELATIVE_HUMIDITY_%`))
# awn.temp.rh     <- subset(awn.temp.rh.na, AIR_TEMP_F < 500 & `RELATIVE_HUMIDITY_%` < 500)
# 
# awn.1h.temp     <- aggregate(awn.temp.rh["AIR_TEMP_F"], list(awn.temp.rh$UNIT_ID, awn.temp.rh$LATITUDE, awn.temp.rh$LONGITUDE, hour=cut(as.POSIXct(awn.temp.rh$TSTAMP_PST)-1, "hour")), mean)
# awn.1h.rh       <- aggregate(awn.temp.rh["RELATIVE_HUMIDITY_%"], list(awn.temp.rh$UNIT_ID, hour=cut(as.POSIXct(awn.temp.rh$TSTAMP_PST)-1, "hour")), mean)
# 
# awn.1h          <- merge(awn.1h.temp, awn.1h.rh, by = c("Group.1","hour"))
# awn.1h$datetime_PST <- ymd_hms(awn.1h$hour, tz="America/Los_Angeles")
# awn.1h$year     <- year(awn.1h$datetime_PST)
# 
# awn.1h.summary  <- ddply(awn.1h, .(Group.1,year), summarize,  TEMP_F=mean(AIR_TEMP_F), RH=mean(`RELATIVE_HUMIDITY_%`), N=length(year))
# 
# setDT(awn.1h)
# 
# awn.1h[, hi := heat_index(AIR_TEMP_F, `RELATIVE_HUMIDITY_%`)]
# 
# awn.1h <- awn.1h %>% rename(Site.Num.awn=Group.1, Latitude.awn=Group.2, Longitude.awn=Group.3)
# 
# rm(awn, awn.1h.rh, awn.1h.temp, awn.temp.rh, awn.temp.rh.na)
# 
# setDF(awn.1h); unique(awn.1h[c("Latitude.awn", "Longitude.awn")])
# 
# save(awn.1h, file =         "Data/awn.1h.2010.2018.RData")
# save(awn.1h.summary, file = "Data/awn.1h.summary.2010.2018.RData")

```

```{r upload-clean-epa-data, include=F, cache=T}

#https://aqs.epa.gov/aqsweb/airdata/download_files.html

# epa.2010 <- read.csv("Data/epa/hourly_88101_2010.csv"); epa.2010.wa <- subset(epa.2010, State.Code=='53')
# epa.2011 <- read.csv("Data/epa/hourly_88101_2011.csv"); epa.2011.wa <- subset(epa.2011, State.Code=='53')
# epa.2012 <- read.csv("Data/epa/hourly_88101_2012.csv"); epa.2012.wa <- subset(epa.2012, State.Code=='53')
# epa.2013 <- read.csv("Data/epa/hourly_88101_2013.csv"); epa.2013.wa <- subset(epa.2013, State.Code=='53')
# epa.2014 <- read.csv("Data/epa/hourly_88101_2014.csv"); epa.2014.wa <- subset(epa.2014, State.Code=='53')
# epa.2015 <- read.csv("Data/epa/hourly_88101_2015.csv"); epa.2015.wa <- subset(epa.2015, State.Code=='53')
# epa.2016 <- read.csv("Data/epa/hourly_88101_2016.csv"); epa.2016.wa <- subset(epa.2016, State.Code=='53')
# epa.2017 <- read.csv("Data/epa/hourly_88101_2017.csv"); epa.2017.wa <- subset(epa.2017, State.Code=='53')
# epa.2018 <- read.csv("Data/epa/hourly_88101_2018.csv"); epa.2018.wa <- subset(epa.2018, State.Code=='53')
# 
# epa <- bind_rows(epa.2010.wa, epa.2011.wa, epa.2012.wa, epa.2013.wa, epa.2014.wa, epa.2015.wa, epa.2016.wa, epa.2017.wa, epa.2018.wa)
# 
# epa$datetime_UTC = ymd_hm(paste(epa$Date.GMT, epa$Time.GMT))
# epa$datetime_PST = as_datetime(epa$datetime, tz = "America/Los_Angeles")
# 
# epa$year     <- year(epa$datetime_PST)
# 
# epa.1h <- subset(epa, select = c(State.Code, County.Code, County.Name, Site.Num, Latitude, Longitude, Datum, Parameter.Name, datetime_PST, year, Date.Local, Time.Local, Date.GMT, Time.GMT, Sample.Measurement, Units.of.Measure, State.Name, County.Name))
# 
# table(epa.1h$County.Name,epa.1h$Site.Num)
# 
# epa.1h <- epa.1h %>% unite("Site", County.Name:Site.Num, sep = "-", remove = FALSE)
# 
# epa.1h.summary  <- ddply(epa.1h, .(Site,year), summarize,  PM=mean(Sample.Measurement), N=length(year))
# 
# setDT(epa.1h)
# 
# rm(epa, epa.2010, epa.2011, epa.2012, epa.2013, epa.2014, epa.2015, epa.2016, epa.2017, epa.2018)
# rm(epa.2010.wa, epa.2011.wa, epa.2012.wa, epa.2013.wa, epa.2014.wa, epa.2015.wa, epa.2016.wa, epa.2017.wa, epa.2018.wa)
# #rm(epa.nearestAWN.merge, locationsAWN, locationsPM, nearestAWN, set1sp, set2sp)
# 
# setDF(epa.1h); unique(epa.1h[c("Latitude", "Longitude")]); unique(epa.1h$Site)
# 
# save(epa.1h, file =         "Data/epa.1h.2010.2018.RData")
# save(epa.1h.summary, file = "Data/epa.1h.summary.2010.2018.RData")

```

```{r merge-awn-epa-data, include=F, cache=T}

# load("Data/epa.1h.summary.2010.2018.RData")
# load("Data/awn.1h.summary.2010.2018.RData")
# load("Data/epa.1h.2010.2018.RData")
# load("Data/awn.1h.2010.2018.RData")
# 
# #find nearest weather station
# locationsPM  = unique(epa.1h[, c("Longitude","Latitude")])
# locationsAWN = unique(awn.1h[, c("Longitude.awn","Latitude.awn")])
# 
# set1sp <- SpatialPoints(locationsPM)
# set2sp <- SpatialPoints(locationsAWN)
# 
# locationsPM$nearest_in_set2 <- apply(gDistance(set2sp, set1sp, byid=T), 1, which.min)
# 
# nearestAWN = locationsAWN[locationsPM$nearest_in_set2,]
# nearestAWN$Longitude.pm = locationsPM$Longitude
# nearestAWN$Latitude.pm = locationsPM$Latitude
# 
# #calculate haversine distance. Returns meters. Assumes radius of the earth is 6378137 m 
# nearestAWN$dist_m = distHaversine((nearestAWN[,c("Longitude.awn", "Latitude.awn")]), (nearestAWN[,c("Longitude.pm", "Latitude.pm")]))
# nearestAWN$dist_mi = nearestAWN$dist_m/1609.34
# 
# #Merge AWN and EPA data
# 
# setDT(nearestAWN)
# 
# epa.nearestAWN.merge = merge(epa.1h, nearestAWN, by.x=c("Longitude","Latitude"), by.y=c("Longitude.pm","Latitude.pm"))
# 
# setDT(epa.nearestAWN.merge); setkey (epa.nearestAWN.merge, Longitude, Latitude, datetime_PST)
# setDT(awn.1h); setkey (awn.1h[!is.na(Longitude.awn)], Longitude.awn, Latitude.awn, datetime_PST)
# 
# epa.awn.final.wide = merge(epa.nearestAWN.merge, awn.1h, by.x= c("Longitude.awn", "Latitude.awn", "datetime_PST"), by.y = c("Longitude.awn", "Latitude.awn", "datetime_PST"))
# 
# epa.awn.final.long <- melt(data = epa.awn.final.wide, id.vars = c("Latitude","Longitude","datetime_PST","Site"), measure.vars = c("Sample.Measurement", "hi"))
# 
# setDF(epa.awn.final.long); z<- unique(epa.awn.final.long[c("Latitude", "Longitude")])
# 
# save(epa.awn.final.wide, file =         "Data/epa.awn.final.wide.2010.2018.RData")
# save(epa.awn.final.long, file =         "Data/epa.awn.final.long.2010.2018.RData")

```

## METHODS

### Heat

- Requested AWN data (with data QA date pre-2010) from locations nearest EPA PM sites for 2010-2018
- Computed heat index using Rothfusz approach

### PM2.5

- Downloaded [EPA PM data for 2010-2018](#https://aqs.epa.gov/aqsweb/airdata/download_files.html)
- Merged with heat data
- Computed nearest Haversine distance between AWN and EPA sites

### Plots

- Histograms for temp, relative humidity, heat index, and PM2.5 (PM2.5 FRM/FEM Mass (88101))
- Density plots
- Time series
- Scatter plot of heat index vs. PM, 2010-2018 by location (n=24)

### Maps

- Ag worker population by county
- PM * HI by county

## RESULTS

```{r summary.tables}

load("Data/epa.awn.final.wide.2010.2018.RData")
load("Data/epa.awn.final.long.2010.2018.RData")
n.am.asd.gm.gsd(epa.awn.final.wide$Sample.Measurement)
n.am.asd.gm.gsd(epa.awn.final.wide$hi)

```

```{r histograms.temp.rh.hi}

ggplot(epa.awn.final.wide, aes(x=AIR_TEMP_F))            + geom_histogram() + theme_HEI(10) + xlab("Air Temperature (F)") + ylab("Count")
ggplot(epa.awn.final.wide, aes(x=`RELATIVE_HUMIDITY_%`)) + geom_histogram() + theme_HEI(10) + xlab("Relative Humidity (%)") + ylab("Count")
ggplot(epa.awn.final.wide, aes(x=hi))                    + geom_histogram() + theme_HEI(10) + xlab("Heat Index") + ylab("Count")
ggplot(epa.awn.final.wide, aes(x=Sample.Measurement))    + geom_histogram() + theme_HEI(10) + xlab("PM2.5 µg/m3") + ylab("Count")

```

```{r density-plots}

ggplot(epa.awn.final.wide[!is.na(hi)], aes( `hi`, y=..density..)) + 
  geom_density() + 
  geom_vline(xintercept = 80, col = "red") +
  facet_wrap(~Site, scales = "free") +
  geom_histogram(aes(y=..density..), alpha=0.5, position="identity") +
  xlim(0, 100) + 
  theme_HEI(12) + 
  xlab("Heat Index")

ggplot(epa.awn.final.wide[!is.na(`Sample.Measurement`)], aes( `Sample.Measurement`, y=..density..)) + 
  geom_density() + 
  geom_vline(xintercept = 35, col = "red") +
  facet_wrap(~Site, scales = "free") +
  geom_histogram(aes(y=..density..), alpha=0.5, position="identity")+
  xlim(0, 100) + 
  theme_HEI(12) + 
  xlab("PM2.5 µg/m3")

```

```{r time.series}

ggplot(epa.awn.final.long, aes(x=datetime_PST, y=value, color=factor(variable))) + geom_line() + facet_wrap(facets = vars(Site)) + theme_HEI(12) + xlab("Year") + ylab("")


```

```{r scatter.plots}
#All data

ggplot(epa.awn.final.wide, aes(x=hi, y=Sample.Measurement)) + geom_point(size=1) + facet_wrap(facets = vars(Site)) + theme_HEI(12) + xlab("Heat Index") + ylab("PM2.5 µg/m3")

#PM data above 35 ug/m3

ggplot(epa.awn.final.wide[epa.awn.final.wide$Sample.Measurement>35], aes(x=hi, y=Sample.Measurement)) + geom_point(size=1) + facet_wrap(facets = vars(Site)) + theme_HEI(12) + xlab("Heat Index") + ylab("PM2.5 µg/m3")

#Heat index data above 80
#


ggplot(epa.awn.final.wide[epa.awn.final.wide$hi>80], aes(x=hi, y=Sample.Measurement)) + geom_point(size=1) + facet_wrap(facets = vars(Site)) + theme_HEI(12) + xlab("Heat Index") + ylab("PM2.5 µg/m3")

#density plot
p_load(viridis)
df <- fortify(epa.awn.final.wide[`Sample.Measurement`>=35 &
                          `hi`>=80])

ggplot(df[Site != "Clark-23"]) +
  geom_hex(bins = 12, aes(x    = `Sample.Measurement`,
               y    = hi,
               fill = ..count..),
           color = "black") +
  scale_fill_viridis_c(option = "inferno") +
  labs(fill = 'Count') +
  facet_wrap(~Site) +
  theme_HEI(12) +
  xlim(0,200) +
  geom_hline(yintercept = 80, col = "red")+
  geom_vline(xintercept = 35, col = "red")



```

## REFERENCES
